---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
```

# blueterra

[![R-CMD-check](https://github.com/el-cordero/blueterra/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/el-cordero/blueterra/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/el-cordero/blueterra/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/el-cordero/blueterra/actions/workflows/pkgdown.yaml)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://github.com/el-cordero/blueterra/blob/main/LICENSE.md)

`blueterra` is an R package for geomorphometric analysis of submerged
terrain. It works from user-supplied bathymetric or elevation rasters and
provides workflows for deriving terrain metrics, organizing metrics into
process-oriented groups, and summarizing seafloor structure across polygons,
transects, depth bands, and isobath corridors.

For a complete worked example using the installed example rasters, see the
[Get Started article](https://el-cordero.github.io/blueterra/articles/blueterra.html).

Full documentation and articles are available at
<https://el-cordero.github.io/blueterra/>.

## Installation

Install the development version from GitHub:

```r
install.packages("remotes")
remotes::install_github("el-cordero/blueterra")
```

From a local source checkout:

```r
install.packages("path/to/blueterra", repos = NULL, type = "source")
```

## Example Data

The installed examples are reduced from analysis rasters and sampling
rectangles used to test terrain workflows on real shelf-margin morphology.
They are compact enough for package examples, but they retain depth gradients,
slope breaks, local relief, and sampling-rectangle geometry.

```{r setup, message = FALSE}
library(blueterra)
library(terra)

hitw <- read_bathy(blueterra_example("hitw"))
hoyo <- read_bathy(blueterra_example("hoyo"))
slope <- read_bathy(blueterra_example("slope"))
rectangles <- terra::vect(blueterra_example("sampling_rectangles"))

hitw_rect <- rectangles[rectangles$site_id == "hitw", ]
```

```{r examples}
examples <- blueterra_examples()
examples$path <- basename(examples$path)
examples
```

## Quick Start

This compact workflow reads Hole-in-the-Wall bathymetry, checks the raster
assumptions, prepares the surface, derives a focused terrain stack, and
summarizes metrics inside the sampling rectangle.

```{r quick-start}
bathy_info(hitw)

hitw_prepared <- prepare_bathy(
  hitw,
  depth_range = c(-220, -25),
  smooth = TRUE,
  smooth_window = 3
)

hitw_metrics <- derive_terrain(
  hitw_prepared,
  metrics = c("slope", "aspect", "northness", "eastness", "tri", "rugosity",
              "bpi", "curvature", "surface_area_ratio")
)

terrain_summary <- summarize_terrain(
  hitw_metrics,
  hitw_rect,
  fun = c("mean", "sd", "min", "max")
)

names(hitw_metrics)
terrain_summary[, c("site_id", "site_name", "slope_deg_mean", "bpi_3x3_mean")]
```

Depth sign conventions are preserved unless conversion is requested
explicitly. The example rasters are stored as negative elevation, so larger
numeric values are shallower and smaller values are deeper.

Square BPI windows are expressed in cells and include the focal cell. Annular
BPI radii are expressed in map units, require a projected CRS, and use each
raster axis's own cell dimension. BPI and VRM-style rugosity use available
focal support at raster edges and along missing-data boundaries; the outermost
derivative cells can still be missing. Normalized BPI returns `NA` when its
focal support has zero variance or no usable values.

For boundary-sensitive polygon summaries, set `exact = TRUE` when the optional
`exactextractr` dependency is available. This uses raster--polygon coverage
fractions to weight means, population standard deviations, medians, sums, and
effective cell counts; minima and maxima use positively intersected cells.

## Key Figures

Hillshade is used in these figures as visual relief. It helps the reader see
the terrain form behind contours, vectors, and metric layers; it is not a
model predictor unless the analyst chooses to include it.

```{r slope-map, eval = requireNamespace("ggplot2", quietly = TRUE), fig.alt = "Slope-clip bathymetry with hillshade, contours, and sampling rectangles."}
plot_bathy(
  slope,
  contours = TRUE,
  contour_interval = 25,
  vectors = rectangles,
  title = "Slope-Clip Bathymetry",
  subtitle = "Hillshade, contours, and sampling rectangles"
)
```

```{r metric-map, eval = requireNamespace("ggplot2", quietly = TRUE), fig.alt = "Slope metric over hillshaded Hole-in-the-Wall bathymetry."}
plot_metric(
  hitw_metrics,
  metric = "slope_deg",
  bathy = hitw_prepared,
  contours = TRUE,
  contour_interval = 25,
  vectors = hitw_rect,
  title = "Slope Over Hillshaded Bathymetry",
  legend_title = "Slope (degrees)"
)
```

```{r profile-and-corridors, eval = requireNamespace("ggplot2", quietly = TRUE), fig.alt = "Bathymetric cross-sections oriented from shallow terrain toward deeper terrain."}
transects <- make_transects(hitw_rect, spacing = 75, bathy = hitw_prepared)
cross_sections <- sample_transects(transects, hitw_prepared, n = 12)

plot_cross_sections(
  cross_sections,
  value_col = "bathy_m",
  mean_profile = TRUE,
  mean_profile_na_rm = TRUE,
  normalize_distance = FALSE,
  profile_direction = "top_to_bottom",
  title = "Bathymetric Cross-Sections",
  subtitle = "Profiles read from shallow to deep terrain"
)
```

Surface-derived transects record `orientation_resultant_length` alongside the
angle and source. Values near one indicate aligned aspect vectors; values near
zero indicate cancelling aspects and an unreliable mean direction, in which
case a manual or bounding-box orientation is more defensible.

```{r isobath-map, eval = requireNamespace("ggplot2", quietly = TRUE), fig.alt = "Isobath corridors over hillshaded bathymetry with source isobaths."}
isobaths <- extract_isobaths(hitw_prepared, depths = c(-50, -80, -120))
corridors <- make_isobath_corridors(
  hitw_prepared,
  depths = c(-50, -80, -120),
  width = 5
)

plot_isobath_corridors(
  corridors,
  hitw_prepared,
  isobaths = isobaths,
  background_contours = FALSE,
  title = "Isobath Corridors",
  subtitle = "Black lines show source isobaths; 5 m is the one-sided buffer distance"
)
```

Here `width = 5` creates a nominal 10 m full-width corridor around each source
isobath. Corridors are independent buffers and may overlap, so their summaries
are not mutually exclusive or additive. The returned features record
`buffer_distance`, `nominal_corridor_width`, and `overlap_policy`.

## Articles

The pkgdown articles carry the full worked examples:

- [Get started](https://el-cordero.github.io/blueterra/articles/blueterra.html)
- [User-supplied rasters](https://el-cordero.github.io/blueterra/articles/user-supplied-rasters.html)
- [Terrain metrics](https://el-cordero.github.io/blueterra/articles/terrain-metrics.html)
- [Process groups](https://el-cordero.github.io/blueterra/articles/process-groups.html)
- [Transects and cross-sections](https://el-cordero.github.io/blueterra/articles/transects-cross-sections.html)
- [Isobath corridors](https://el-cordero.github.io/blueterra/articles/isobath-corridors.html)
- [Custom metrics and process groups](https://el-cordero.github.io/blueterra/articles/custom-metrics-process-groups.html)
- [Visual proof](https://el-cordero.github.io/blueterra/articles/visual-proof.html)

## Citation

Please cite `blueterra` with the package citation once the release metadata are
finalized:

```{r citation, eval = FALSE}
citation("blueterra")
```

## License

`blueterra` is released under the MIT license.
