---
title: "blueterra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blueterra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
library(terra)
```

`blueterra` supports submerged-terrain geomorphometry from bathymetric and
elevation rasters. The package keeps raster and vector operations in `terra`,
which helps keep CRS, extent, resolution, and geometry handling consistent
throughout an analysis.

## Start with Real Example Data

```{r read}
hitw <- read_bathy(blueterra_example("hitw"))
rectangles <- terra::vect(blueterra_example("sampling_rectangles"))
hitw_rect <- rectangles[rectangles$site_id == "hitw", ]

bathy_info(hitw)
hitw_rect
```

## Prepare the Surface

```{r prepare}
prepared <- prepare_bathy(
  hitw,
  depth_range = c(-220, -25),
  smooth = TRUE,
  smooth_window = 3
)
check_bathy_units(prepared, units = "m", positive_depth = FALSE)
```

Depth signs are preserved unless a conversion is explicit:

```{r signs}
positive_depth <- set_depth_positive(prepared)
range(terra::values(positive_depth), na.rm = TRUE)
```

## Derive Terrain Metrics

```{r metrics}
terrain <- derive_terrain(
  prepared,
  metrics = c("slope", "aspect", "northness", "eastness", "tri", "bpi",
              "curvature")
)
names(terrain)
```

## Summarize Sampling Rectangles

```{r zones}
summarize_terrain(
  terrain,
  hitw_rect,
  fun = c("mean", "sd", "min", "max")
)
```

## Model-Ready Tables

```{r model-ready}
cells <- sample_terrain_cells(terrain, size = 40, method = "regular")
terrain_correlation(cells[, c("slope_deg", "tri", "bpi_3x3", "curvature")])
terrain_pca(cells[, c("slope_deg", "tri", "bpi_3x3", "curvature")])$variance
```
