---
title: "blueterra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blueterra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
```

`blueterra` is for submerged-terrain geomorphometry after the user already has a
raster. It accepts local bathymetric or elevation rasters through `terra` and
does not download, discover, mosaic, or cache BlueTopo data.

## Start with a Raster

```{r read}
bathy <- read_bathy(blueterra_example("bathy"))
bathy_info(bathy)
```

The example raster is synthetic and small enough for examples and tests.

## Prepare the Surface

```{r prepare}
prepared <- prepare_bathy(
  bathy,
  depth_range = c(-90, -20),
  smooth = TRUE,
  smooth_window = 3
)
```

Depth signs are preserved unless a conversion is explicit:

```{r signs}
positive_depth <- set_depth_positive(prepared)
negative_depth <- set_depth_negative(positive_depth)
check_bathy_units(negative_depth, units = "m", positive_depth = FALSE)
```

## Derive Terrain Metrics

```{r metrics}
terrain <- derive_terrain(
  prepared,
  metrics = c("slope", "aspect", "northness", "eastness", "bpi", "roughness")
)
names(terrain)
```

## Summarize Zones

```{r zones}
sites <- sf::st_read(blueterra_example("sites"), quiet = TRUE)
summarize_terrain(terrain, sites)
```

## Model-Ready Tables

```{r model-ready}
cells <- sample_terrain_cells(terrain, size = 25)
terrain_correlation(cells)
```
