---
title: "Isobath Corridors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Isobath Corridors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
```

Isobath corridors summarize terrain near depth contours. They are useful when
observations, surveys, or interpretations follow depth horizons rather than
rectangular or administrative boundaries.

```{r data}
bathy <- read_bathy(blueterra_example("hitw"))
prepared <- prepare_bathy(bathy, depth_range = c(-220, -25), smooth = TRUE)
terrain <- derive_terrain(prepared, metrics = c("slope", "bpi", "curvature"))
```

The example bathymetry stores depth as negative elevation, so contour depths are
passed as negative values:

```{r isobaths}
isobaths <- extract_isobaths(prepared, depths = c(-50, -80, -120))
isobaths
```

Buffers are measured in map units, so the raster should use a projected CRS:

```{r corridors}
corridors <- make_isobath_corridors(
  prepared,
  depths = c(-50, -80, -120),
  width = 20
)
summarize_isobath_terrain(terrain, corridors)
```
