---
title: "Process Groups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Process Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
```

`blueterra` groups terrain metrics into process-oriented terrain-form groups.
The groups support interpretation, variable selection, and reporting. They are
terrain-form proxies; they do not directly measure currents, sediment flux, or
ecological processes.

```{r catalog}
metric_catalog()
```

## Assigning Groups

```{r assign}
bathy <- read_bathy(blueterra_example("hitw"))
terrain <- derive_terrain(
  bathy,
  metrics = c("slope", "aspect", "northness", "eastness", "tri", "bpi",
              "curvature")
)
assign_process_groups(terrain)
summarize_process_groups(terrain)
```

## Representatives

Representative metrics are starting points for compact reporting. The final
choice should still follow the resolution, sampling design, and feature scale
of the analysis.

```{r reps}
select_process_representatives(metrics_available = names(terrain))
```
