---
title: "Terrain Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Terrain Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
```

Terrain derivatives are scale-sensitive. Resolution, smoothing, CRS, and focal
window size affect slope, aspect, BPI, TPI, curvature, rugosity, and surface
area style metrics.

```{r metrics}
bathy <- read_bathy(blueterra_example("bathy"))
terrain <- derive_metric_stack(
  bathy,
  metrics = c("slope", "aspect", "northness", "eastness", "bpi", "roughness",
              "tri", "tpi", "curvature", "surface_area_ratio")
)
names(terrain)
```

## Orientation

Aspect is circular. Northness and eastness convert aspect into linear
components:

```{r orientation}
orientation <- derive_metric_stack(
  bathy,
  metrics = c("aspect", "northness", "eastness")
)
names(orientation)
```

## BPI and Scale

Bathymetric position index is a focal metric. Window size changes the scale of
terrain position being summarized:

```{r bpi}
derive_multiscale_bpi(bathy, windows = c(3, 7, 11))
```

For elevation-like negative bathymetry, positive BPI means higher stored values
relative to the neighborhood. For positive-depth rasters, interpretation is
reversed unless users convert the sign convention first.
