---
title: "Terrain Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Terrain Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
```

Terrain derivatives are scale-sensitive. Resolution, smoothing, CRS, and focal
window size affect slope, aspect, BPI, TPI, curvature, rugosity, and
surface-area style metrics. The examples use reduced Hole-in-the-Wall
bathymetry stored as negative elevation in metres.

```{r data}
bathy <- read_bathy(blueterra_example("hitw"))
prepared <- prepare_bathy(bathy, depth_range = c(-220, -25), smooth = TRUE)
```

## Slope, Aspect, and Orientation

```{r orientation}
orientation <- derive_metric_stack(
  prepared,
  metrics = c("slope", "aspect", "northness", "eastness")
)
names(orientation)
```

Aspect is circular. Northness and eastness convert aspect into linear
components that can be summarized in tables or used as model predictors.

## Roughness, TRI, Rugosity, and Surface Area

```{r structure}
structure_metrics <- derive_metric_stack(
  prepared,
  metrics = c("roughness", "tri", "rugosity", "surface_area_ratio")
)
names(structure_metrics)
```

Rugosity is computed as vector ruggedness over a focal window. Surface-area
ratio is derived from slope and should be interpreted as a local raster-cell
surface approximation, not as a measured benthic surface area.

## BPI, TPI, and Scale

```{r bpi}
fine_bpi <- derive_bpi(prepared, window = 3)
broad_bpi <- derive_bpi(prepared, window = 11)
multi_bpi <- derive_multiscale_bpi(prepared, windows = c(3, 7, 11))

names(multi_bpi)
terra::global(fine_bpi, c("min", "mean", "max"), na.rm = TRUE)
terra::global(broad_bpi, c("min", "mean", "max"), na.rm = TRUE)
```

For elevation-like negative bathymetry, positive BPI means a cell has a higher
stored value than its neighborhood. For positive-depth rasters, interpretation
is reversed unless the sign convention is converted first.

## Curvature

```{r curvature}
curvature <- derive_curvature(prepared)
terra::global(curvature, c("min", "mean", "max"), na.rm = TRUE)
```

The curvature layer is a Laplacian-style local index based on a four-neighbor
kernel. It is useful as a compact measure of local convexity or concavity, but
it is not profile curvature or plan curvature.
