---
title: "Terrain metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Terrain metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup, message = FALSE}
library(blueterra)
library(terra)
```

Terrain derivatives are scale-sensitive. Resolution, smoothing, CRS, and focal
window size affect slope, aspect, BPI, TPI, curvature, rugosity, and
surface-area style metrics. The examples use reduced Hole-in-the-Wall
bathymetry from the southwest Puerto Rico shelf margin near La Parguera,
Puerto Rico.

```{r data}
bathy <- read_bathy(blueterra_example("hitw"))
prepared <- prepare_bathy(bathy, depth_range = c(-220, -25), smooth = TRUE)
```

## Slope, Aspect, and Orientation

Slope and aspect describe local orientation. Aspect is circular, so northness
and eastness convert it into two linear components that can be summarized in
tables or used as predictors.

```{r orientation}
orientation <- derive_metric_stack(
  prepared,
  metrics = c("slope", "aspect", "northness", "eastness")
)
names(orientation)
terra::global(orientation[["slope_deg"]], c("min", "mean", "max"), na.rm = TRUE)
```

```{r orientation-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Slope over hillshaded bathymetry."}
plot_metric(
  orientation,
  "slope_deg",
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Slope Over Hillshade"
)
```

## Roughness, TRI, Rugosity, and Surface Area

These metrics describe local relief variability. Their values change with grid
resolution and focal-window size, so windows should be chosen to match the
feature scale being interpreted.

The VRM-style rugosity calculation averages available slope/aspect-derived
normal vectors within its focal window. It can therefore use partial focal
support beside raster edges or missing-data boundaries, although the outermost
cells can remain missing because slope and aspect require neighbouring
elevation cells.

```{r structure}
structure_metrics <- derive_metric_stack(
  prepared,
  metrics = c("roughness", "tri", "rugosity", "surface_area_ratio")
)
names(structure_metrics)
terra::global(structure_metrics[["tri"]], c("min", "mean", "max"), na.rm = TRUE)
terra::global(structure_metrics[["surface_area_ratio"]], c("min", "mean", "max"), na.rm = TRUE)
```

```{r structure-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Rugosity over hillshaded bathymetry."}
plot_metric(
  structure_metrics,
  "rugosity_vrm_3x3",
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Rugosity Over Hillshade"
)
```

Surface-area ratio is derived from `1 / cos(slope)` and should be interpreted
as a local raster-cell surface approximation, not as a triangulated or measured
benthic surface area.

```{r surface-area-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Surface-area ratio over hillshaded bathymetry."}
plot_metric(
  structure_metrics,
  "surface_area_ratio",
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Surface-Area Ratio Over Hillshade"
)
```

## BPI, TPI, and Scale

BPI and TPI compare a cell with its neighborhood. For elevation-like negative
bathymetry, positive BPI means a cell has a higher stored value than its
neighborhood, which usually means shallower terrain. For positive-depth rasters,
interpretation reverses unless the sign convention is converted first.

The square BPI window is specified in cells and includes the focal cell. An
annular BPI window is specified by inner and outer radii in map units; it
requires a projected CRS and calculates its footprint with separate x- and
y-cell dimensions when the grid is not square. A positive inner radius excludes
the focal cell. BPI uses available values in partial focal support at edges and
missing-data boundaries, while a missing focal value remains missing. With
`normalize = TRUE`, zero-variance or unavailable focal neighbourhoods return
`NA` rather than a numerical `NaN`.

```{r bpi}
fine_bpi <- derive_bpi(prepared, window = 3)
broad_bpi <- derive_bpi(prepared, window = 11)
annular_bpi <- derive_bpi(prepared, inner_radius = 8, outer_radius = 24)
normalized_bpi <- derive_bpi(prepared, window = 3, normalize = TRUE)
tpi <- derive_tpi(prepared)
multi_bpi <- derive_multiscale_bpi(prepared, windows = c(3, 7, 11))

names(multi_bpi)
terra::global(fine_bpi, c("min", "mean", "max"), na.rm = TRUE)
terra::global(broad_bpi, c("min", "mean", "max"), na.rm = TRUE)
terra::global(tpi, c("min", "mean", "max"), na.rm = TRUE)
```

```{r bpi-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="BPI over hillshaded bathymetry."}
plot_metric(
  fine_bpi,
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Fine-Scale BPI Over Hillshade"
)
```

## Curvature

```{r curvature}
curvature <- derive_curvature(prepared)
terra::global(curvature, c("min", "mean", "max"), na.rm = TRUE)
```

The curvature layer is a Laplacian-style local index based on a four-neighbor
kernel. It is useful as a compact measure of local convexity or concavity, but
it is not profile curvature or plan curvature.

```{r curvature-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Curvature over hillshaded bathymetry."}
plot_metric(
  curvature,
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Local Curvature Over Hillshade"
)
```

## Methodological Background and Related Software

Lindsay (2016, <https://doi.org/10.1016/j.cageo.2016.07.003>) provides a
concise geomorphometric software and terrain-analysis reference through the
Whitebox GAT case study. The R package `whitebox` is related software that
provides an R frontend to WhiteboxTools: Wu, Q. and Brown, A. (2025).
`whitebox`: WhiteboxTools R Frontend. R package version 2.4.3.
<https://doi.org/10.32614/CRAN.package.whitebox>.

These references are included as methodological background and software context
for terrain analysis. The examples in this vignette use `terra` operations
directly.
