---
title: "Terrain Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Terrain Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup, message = FALSE}
library(blueterra)
library(terra)
```

Terrain derivatives are scale-sensitive. Resolution, smoothing, CRS, and focal
window size affect slope, aspect, BPI, TPI, curvature, rugosity, and
surface-area style metrics. The examples use reduced Hole-in-the-Wall
bathymetry from the southwest Puerto Rico shelf margin near La Parguera,
Puerto Rico.

```{r data}
bathy <- read_bathy(blueterra_example("hitw"))
prepared <- prepare_bathy(bathy, depth_range = c(-220, -25), smooth = TRUE)
```

## Slope, Aspect, and Orientation

Slope and aspect describe local orientation. Aspect is circular, so northness
and eastness convert it into two linear components that can be summarized in
tables or used as predictors.

```{r orientation}
orientation <- derive_metric_stack(
  prepared,
  metrics = c("slope", "aspect", "northness", "eastness")
)
names(orientation)
terra::global(orientation[["slope_deg"]], c("min", "mean", "max"), na.rm = TRUE)
```

```{r orientation-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Slope over hillshaded bathymetry."}
plot_metric(
  orientation,
  "slope_deg",
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Slope Over Hillshade"
)
```

## Roughness, TRI, Rugosity, and Surface Area

These metrics describe local relief variability. Their values change with grid
resolution and focal-window size, so windows should be chosen to match the
feature scale being interpreted.

```{r structure}
structure_metrics <- derive_metric_stack(
  prepared,
  metrics = c("roughness", "tri", "rugosity", "surface_area_ratio")
)
names(structure_metrics)
terra::global(structure_metrics[["tri"]], c("min", "mean", "max"), na.rm = TRUE)
```

```{r structure-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Rugosity over hillshaded bathymetry."}
plot_metric(
  structure_metrics,
  "rugosity_vrm_3x3",
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Rugosity Over Hillshade"
)
```

Surface-area ratio is derived from slope and should be interpreted as a local
raster-cell surface approximation, not as a measured benthic surface area.

## BPI, TPI, and Scale

BPI and TPI compare a cell with its neighborhood. For elevation-like negative
bathymetry, positive BPI means a cell has a higher stored value than its
neighborhood, which usually means shallower terrain. For positive-depth rasters,
interpretation reverses unless the sign convention is converted first.

```{r bpi}
fine_bpi <- derive_bpi(prepared, window = 3)
broad_bpi <- derive_bpi(prepared, window = 11)
multi_bpi <- derive_multiscale_bpi(prepared, windows = c(3, 7, 11))

names(multi_bpi)
terra::global(fine_bpi, c("min", "mean", "max"), na.rm = TRUE)
terra::global(broad_bpi, c("min", "mean", "max"), na.rm = TRUE)
```

```{r bpi-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="BPI over hillshaded bathymetry."}
plot_metric(
  fine_bpi,
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Fine-Scale BPI Over Hillshade"
)
```

## Curvature

```{r curvature}
curvature <- derive_curvature(prepared)
terra::global(curvature, c("min", "mean", "max"), na.rm = TRUE)
```

The curvature layer is a Laplacian-style local index based on a four-neighbor
kernel. It is useful as a compact measure of local convexity or concavity, but
it is not profile curvature or plan curvature.

```{r curvature-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Curvature over hillshaded bathymetry."}
plot_metric(
  curvature,
  bathy = prepared,
  contours = TRUE,
  contour_interval = 25,
  title = "Local Curvature Over Hillshade"
)
```
