---
title: "Transects and cross-sections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transects and cross-sections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup, message = FALSE}
library(blueterra)
library(terra)
```

Transects turn a bathymetric raster into cross-sectional profiles. They are
useful for showing terraces, escarpments, slope breaks, and cross-shelf
gradients that are not always obvious from cell-by-cell terrain metrics.

## Data and Surface Orientation

```{r data}
bathy <- read_bathy(blueterra_example("hitw"))
rectangles <- terra::vect(blueterra_example("sampling_rectangles"))
hitw_rect <- rectangles[rectangles$site_id == "hitw", ]

prepared <- prepare_bathy(
  bathy,
  depth_range = c(-220, -25),
  smooth = TRUE
)

estimate_surface_orientation(prepared, hitw_rect)
```

`estimate_surface_orientation()` converts local aspect into a representative
bearing and a transect angle. Slope weighting emphasizes cells where orientation
is better defined.

## Automatic Transect Orientation

```{r automatic}
transects <- make_transects(
  hitw_rect,
  spacing = 75,
  bathy = prepared
)

transects[, c("transect_id", "angle_deg", "angle_source", "mean_aspect_deg")]
```

```{r automatic-map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Automatically oriented transects over hillshaded bathymetry."}
plot_transects(
  prepared,
  transects,
  color_by = "transect_id",
  show_legend = FALSE,
  contour_interval = 25,
  title = "Terrain-Oriented Transects",
  subtitle = "Line angle estimated from local surface aspect"
)
```

## Manual Angle Override

Manual angles remain useful when transects need to match survey design, a
predefined grid, or a known field orientation.

```{r manual}
manual_transects <- make_transects(
  hitw_rect,
  spacing = 75,
  angle = 45
)

manual_transects[, c("transect_id", "angle_deg", "angle_source")]
```

## Sampling Cross-Sections

`sample_transects()` extracts raster values at regular positions along each
line. `extract_cross_sections()` is an alias for the same operation when the
output is being used as a profile table.

```{r sample}
samples <- sample_transects(transects, prepared, n = 12)
sections <- extract_cross_sections(transects, prepared, n = 12)

head(samples[, c("transect_id", "distance", "x", "y", "bathy_m")])
summarize_cross_sections(samples, value_col = "bathy_m")
identical(names(samples), names(sections))
```

```{r cross-section-plot, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Bathymetric cross-sections with bathy_m on the y-axis."}
plot_cross_sections(
  samples,
  value_col = "bathy_m",
  show_legend = TRUE,
  mean_profile = TRUE,
  normalize_distance = TRUE,
  title = "Bathymetric Cross-Sections"
)
```

The value column is explicit. This matters because transect tables include
numeric metadata such as `width_m`, `height_m`, `angle_deg`, and `offset`.

## Single-Transect Profile

```{r profile, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Bathymetric profile along a single transect."}
one <- samples[samples$transect_id == samples$transect_id[1], ]

plot_depth_profile(
  one,
  value_col = "bathy_m",
  title = "Bathymetry Along One Transect"
)
```

## Metric Profiles

The same plotting functions work with derived metrics. Here the profile shows
slope rather than bathymetric elevation.

```{r slope-profile, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="Slope profile along one transect."}
metrics <- derive_terrain(prepared, metrics = c("slope", "tri", "bpi"))
metric_samples <- sample_transects(
  transects,
  metrics[["slope_deg"]],
  n = 25
)
metric_one <- metric_samples[metric_samples$transect_id == metric_samples$transect_id[1], ]

plot_depth_profile(
  metric_one,
  value_col = "slope_deg",
  title = "Slope Along One Transect"
)
```

The profile y-axis changes with the selected raster value. The function name is
historical; the plotted variable can be bathymetry, elevation, slope, rugosity,
BPI, or a custom metric.
