---
title: "Working with user-supplied rasters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with user-supplied rasters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup, message = FALSE}
library(blueterra)
library(terra)
```

`blueterra` starts with a local raster path or an existing
`terra::SpatRaster`. Region, datum, grid resolution, and depth sign convention
stay visible in the analysis code because those choices affect interpretation.

## Raster Paths and SpatRasters

```{r path}
path <- blueterra_example("hoyo")
hoyo <- read_bathy(path)
same_hoyo <- as_bathy(hoyo)

class(path)
class(hoyo)
class(same_hoyo)
bathy_info(hoyo)
```

`read_bathy()` is the local file reader. `as_bathy()` is useful when functions
need to accept either file paths or raster objects.

## Vector Inputs

Use `terra::vect()` for local vector files. Functions that take zones, masks,
or transect boundaries also accept a local vector path.

```{r vector}
rectangles <- terra::vect(blueterra_example("sampling_rectangles"))
rectangles[, c("site_id", "site_name", "feature_type")]

hoyo_rect <- rectangles[rectangles$site_id == "hoyo", ]
masked <- mask_bathy(hoyo, hoyo_rect)
class(masked)
```

## CRS

```{r crs}
check_bathy_crs(hoyo)
terra::crs(hoyo, proj = TRUE)
terra::res(hoyo)
```

Slope, transect spacing, and focal windows in map units should be interpreted
in a projected CRS. In particular, annular BPI radii require a projected CRS
and use separate x- and y-cell dimensions when the grid is not square. Isobath
corridor buffering accepts longitude/latitude input with a warning, but a width
in degrees is usually not an interpretable corridor distance; reproject before
interpreting corridor width. `blueterra` does not silently reproject an input
raster for these operations.

```{r project}
coarse_template <- terra::aggregate(hoyo, fact = 2)
projected <- project_bathy(coarse_template, terra::crs(hoyo))
class(projected)
```

## Depth Convention

Bathymetric rasters are commonly stored as negative elevation or positive
depth. `blueterra` preserves the stored convention unless conversion is
requested explicitly.

```{r sign}
check_bathy_units(hoyo, units = "m", positive_depth = FALSE)
range(terra::values(hoyo), na.rm = TRUE)

positive_depth <- set_depth_positive(hoyo)
range(terra::values(positive_depth), na.rm = TRUE)
```

Depth bands follow the stored values unless `positive_depth = TRUE` is used.

```{r bands}
summarize_depth_bands(
  hoyo,
  breaks = c(-260, -180, -120, -60, -20)
)
```

## Visual Check

The first map should show the measured bathymetric surface and the relief
structure that will influence slope, position, and curvature metrics.

```{r map, eval=requireNamespace("ggplot2", quietly = TRUE), fig.alt="El Hoyo bathymetry with hillshade, contours, and sampling rectangle."}
plot_bathy(
  hoyo,
  contours = TRUE,
  contour_interval = 25,
  vectors = hoyo_rect,
  title = "El Hoyo Bathymetry",
  subtitle = "Hillshade, contours, and sampling rectangle"
)
```
