---
title: "User-Supplied Rasters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-Supplied Rasters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
library(terra)
```

`blueterra` works with digital bathymetric models, digital elevation models, and
other raster surfaces readable by `terra`. A raster can enter the workflow as a
local file path or as an existing `terra::SpatRaster`.

## Raster Paths and SpatRasters

```{r path}
path <- blueterra_example("hoyo")
hoyo <- read_bathy(path)
same_hoyo <- as_bathy(hoyo)

class(path)
class(hoyo)
class(same_hoyo)
```

## Vector Inputs

Use `terra::vect()` for local vector files:

```{r vector}
rectangles <- terra::vect(blueterra_example("sampling_rectangles"))
rectangles[, c("site_id", "site_name", "feature_type")]
```

Functions that take zones, masks, or transect boundaries also accept a local
vector path:

```{r vector-path}
hoyo_rect <- rectangles[rectangles$site_id == "hoyo", ]
masked <- mask_bathy(hoyo, hoyo_rect)
class(masked)
```

## CRS

```{r crs}
check_bathy_crs(hoyo)
```

Slope, buffers, transects, focal windows in map units, and isobath corridors
should be interpreted in a projected CRS. `blueterra` does not silently
reproject; use `project_bathy()` or `terra::project()` when a CRS conversion is
part of the analysis design.

## Depth Convention

```{r sign}
check_bathy_units(hoyo, units = "m", positive_depth = FALSE)
positive_depth <- set_depth_positive(hoyo)
range(terra::values(positive_depth), na.rm = TRUE)
```

Depth bands follow the stored values unless `positive_depth = TRUE` is used.

```{r bands}
summarize_depth_bands(
  hoyo,
  breaks = c(-260, -180, -120, -60, -20)
)
```
