---
title: "User-Supplied Rasters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User-Supplied Rasters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(blueterra)
```

`blueterra` works with user-supplied digital bathymetric models, digital
elevation models, and other raster surfaces readable by `terra`.

The package does not download BlueTopo data. BlueTopo download and mosaicking
are intentionally outside the scope of `blueterra` and will be handled by a
separate package.

## Accepted Inputs

Use a local raster path:

```{r path}
path <- blueterra_example("bathy")
bathy <- read_bathy(path)
```

Or use an existing `terra::SpatRaster`:

```{r spatraster}
same_bathy <- as_bathy(bathy)
```

## CRS

```{r crs}
check_bathy_crs(bathy)
```

Use projected CRS when slope, buffers, focal windows, or distances need linear
map units. `blueterra` warns or errors only where the operation requires a CRS;
it does not silently reproject.

## Depth Convention

Some bathymetry stores depth as negative elevation, while other workflows use
positive depth. `blueterra` preserves signs by default:

```{r sign}
check_bathy_units(bathy, units = "m", positive_depth = FALSE)
set_depth_positive(bathy)
```
